---
title: "DATA VISUALIZATION WITH BASE R"
author: "Sofía Galván and Miranta Kouvari"
date: "2023-08-28"
output: html_document
---

1. First, let's load and prepare the packages and data we are going to work with. In this case, we'll use the `tetrapods` dataset provided by `palaeoverse` package, and we'll explore it.

```{r, message = FALSE, warning= FALSE}
library(palaeoverse)
library(Hmisc)
library(vioplot)

data(tetrapods)
summary(tetrapods$lat) #Summaries of numeric variables
table(tetrapods$order) #Counts of a variable
head(tetrapods)[1:10] #First columns of the dataframe
str(tetrapods) #General structure: 5270 obs and 32 variables
?tetrapods

tetrapods$mid_age <- (tetrapods$max_ma + tetrapods$min_ma)/2 #Roughly calculate medium ages.
```

2. Now, let's explore the distribution of one variable through histograms (for continuous variables) and bar plots (for discrete ones).

```{r, warning = FALSE}
hist(tetrapods$mid_age)
hist(tetrapods$mid_age, breaks=seq(from = 240, to = 360, by = 5), 
     xlab = "Middle age (Ma)",
     main = "")

dens_line <- density(tetrapods$mid_age)
hist(tetrapods$mid_age, breaks=seq(from = 240, to = 360, by = 5),
     xlab = "", main = "Middle age (Ma)", freq = F)
lines(dens_line)

barplot(table(tetrapods$diet))
```

3. When exploring more than one variable at a time, we can use scatter plots.

```{r, warning = FALSE}
plot(x = tetrapods$mid_age, tetrapods$lat)
```

We can add other things to the plot, such as a geological axis with the `axis_geo` function of the `palaeoverse` package:

```{r, warning = FALSE}
plot(x = tetrapods$mid_age, tetrapods$lat, ylab = "Latitude (\u00B0)", 
     xlim = c(350, 250), xaxt = "n", xlab = "Time (Ma)", type = "p", pch = 20)
axis_geo(side = 1, intervals = "periods")
```

And adding more things:

```{r}
my_colors <- c("#f2cb61", "#66B2FF", "#FF5CAD")
plot(x = tetrapods$mid_age, tetrapods$lat, ylab = "Latitude (\u00B0)",
     ylim = c(-90,90), yaxt = "n", xlab = NA, xlim = c(354, 250), xaxt = "n", 
     type = "p", pch = 20, cex = 1.3, col = my_colors[factor(tetrapods$class)])
legend("bottomleft", legend = levels(factor(tetrapods$class)),
       pch = 20, col = my_colors)
axis(2, at = c(-90, -50, 0, 50, 90))
axis_geo(side = 1, intervals = list("epochs", "periods"),
         lab_size = 0.6, exact = T, round = 1)
```

4. If we want to compare between groups, box and violin plots are handy.

```{r}
table(tetrapods$diet)
boxplot(tetrapods$mid_age)

colcajas <- c("#dfcc74", "#51a3a3", "#cb904d", "#75485e")
boxplot(lat ~ diet, data = tetrapods, xlab = "Diet", ylab = "Latitude (\u00B0)",
        ylim = c(-90,90), axes = F, cex.axis = 0.9, boxwex = 0.6, 
        outline = F, col = colcajas)
axis(2, at = c(-90, -50, 0, 50, 90))
axis(1, labels = c("Carnivore", "Herbivore", "Insectivore/Herbivore", "Piscivore"), 
     at = c(1:4))
minor.tick(nx = 1, ny = 2)
```

```{r}
#Violin plot
vioplot(lat ~ diet, data = tetrapods, xlab = "Diet", ylab = "Latitude (\u00B0)",
        ylim = c(-90,90), xaxt = "n", yaxt = "n", axes = F, cex.axis = 0.9, 
        boxwex = 0.6, outline = F, col = colcajas)
axis(2, at = c(-90, -50, 0, 50, 90))
axis(1, labels = c("Carnivore", "Herbivore", "Insectivore/Herbivore", "Piscivore"), 
     at = c(1:4))
minor.tick(nx = 1, ny = 2)
```

5. Lastly, we can save our plots with `pdf`or `svg` function, among others.

```{r, message = FALSE}
svg("DV_baseRplot1.svg", width = 10, height = 7)
vioplot(lat ~ diet, data = tetrapods, xlab = "Diet", ylab = "Latitude (\u00B0)",
        ylim = c(-90,90), xaxt = "n", yaxt = "n", axes = F, cex.axis = 0.9, 
        boxwex = 0.6, outline = F, col = colcajas)
axis(2, at = c(-90, -50, 0, 50, 90))
axis(1, labels = c("Carnivore", "Herbivore", "Insectivore/Herbivore", "Piscivore"), 
     at = c(1:4))
minor.tick(nx = 1, ny = 2)
dev.off()
```

6. Lastly, we should cite the packages we used, for example:

```{r}
citation("palaeoverse") 
```


### References

R for Palaeobiologists: Workshop and Hackathon. https://hackathon.palaeoverse.org/