---
title: "DATA VISUALIZATION WITH GGPLOT2"
author: "Sofía Galván and Miranta Kouvari"
date: "2023-08-25"
output: html_document
---

Before starting, an equivalent script using Base R for plotting will be available on the workshop webpage.

### Let's begin!


1. First, let's load and prepare the packages and data we are going to work with. In this case, we'll use the `tetrapods` dataset provided by `palaeoverse` package. Also, we'll start preparing these data following [Data preparation](https://hackathon.palaeoverse.org/preparation.html) previous tutorial. 

```{r, message = FALSE, warning= FALSE}
library(palaeoverse)
library(ggplot2)
library(svglite)
library(deeptime)
library(extrafont)
data(tetrapods)
?tetrapods

# Assign a common time scale based on an interval key
tetrapods <- look_up(occdf = tetrapods,
                     early_interval = "early_interval",
                     late_interval = "late_interval",
                     int_key = interval_key)

# Making any values which could not be matched contain their original values
tetrapods$interval_max_ma <- ifelse(is.na(tetrapods$interval_max_ma),
                                    tetrapods$max_ma,
                                    tetrapods$interval_max_ma)
tetrapods$interval_min_ma <- ifelse(is.na(tetrapods$interval_min_ma),
                                    tetrapods$min_ma,
                                    tetrapods$interval_min_ma)
tetrapods$interval_mid_ma <- (tetrapods$min_ma + tetrapods$max_ma)/2

#Count number of occurrences per interval
tetrapods_time <- group_apply(occdf = tetrapods, group = "interval_mid_ma", fun = nrow)
tetrapods_time$nrow <- as.numeric(tetrapods_time$nrow)
tetrapods_time$interval_mid_ma <- as.numeric(tetrapods_time$interval_mid_ma)
```

*The general code for plotting with ggplot will be:*

`ggplot(data = my_data, aes(x = my_var1, y = my_var2)) + geom_point()`


2. Now, let's explore the distribution of one variable through histograms (for continuous variables) and bar plots (for discrete ones).

```{r, warning= FALSE}

ggplot(data = tetrapods, mapping = aes(x = interval_mid_ma)) +
  geom_histogram(breaks=seq(from = 240, to = 360, by = 5),
                 color="black", fill="lavender")

ggplot(data = tetrapods, mapping = aes(x = interval_mid_ma)) +
  geom_histogram(aes(y = after_stat(density)),
                 breaks=seq(from = 240, to = 360, by = 5),
                 color="black", fill="lavender") +
  geom_density() #Adding a density curve

ggplot(data = tetrapods, mapping = aes(x =diet)) +
  geom_bar(color="black", fill="lavender")
```


3. When exploring more than one variable at a time, we can use scatter plots.

```{r, warning = FALSE}
ggplot(data = tetrapods, mapping = aes(x = interval_mid_ma, y = lat)) + 
  geom_point() #For a continous variable
ggplot(data = tetrapods, mapping = aes(x = diet, y = lat)) + 
  geom_point() #For a discrete variable
```

We can add other things to the plot, such as a geological axis with the `coor_geo` function of the `deeptime` package:

```{r, warning = FALSE}
ggplot(data = tetrapods, 
       mapping = aes(x = interval_mid_ma, y = lat, color = as.factor(class))) + 
  geom_point() +
  xlim(243,355) +
  xlab("Minimun age (Ma)") + ylab("Latitude (\u00B0)") +
  coord_geo(ylim = c(-90,90))
```

And adding more things:

```{r}
ggplot(data = tetrapods, mapping = aes(x = min_ma, y = lat, color = as.factor(class))) + 
  geom_point(size = 2, shape = 16) +
  xlab("") + ylab("Latitude (\u00B0)") +
  xlim(350, 240) + 
  scale_y_continuous(limits = c(-90, 90), breaks = c(-90,-50,0,50,90)) +
  scale_color_manual(values = c("#f2cb61", "#66B2FF", "#FF5CAD")) +
  theme_classic() + #Here, we're changing the theme of the plot
  theme(legend.title = element_blank()) + #and removing the legend title
  coord_geo(dat = list("epochs", "periods"), pos = list("b", "b"),
            xlim = c(350, 240), ylim = c(-90,90), size = 4.5,
            center_end_labels = T)
```


4. If we want to compare between groups, box and violin plots are handy.

```{r}
ggplot(data = tetrapods, mapping = aes(x = diet, y = lat)) + 
  geom_boxplot()

tetrapods <- tetrapods[!is.na(tetrapods$diet),] #Remove entries with no diet information
colcajas <- c("#dfcc74", "#51a3a3", "#cb904d", "#75485e")

#And adding something more:
ggplot(data = tetrapods, mapping = aes(x = diet, y = lat)) + 
  geom_boxplot(fill = colcajas, outlier.shape = NA, width = 0.6) +
  xlab("Diet") + ylab("Latitude (\u00B0)") +
  scale_y_continuous(limits = c(-90, 90), breaks = c(-90,-50,0,50,90)) +
  scale_x_discrete(labels = c("Carnivore", "Herbivore", "Insectivore/Herbivore", "Piscivore")) +
  theme_classic()
```

```{r}
#We will store this plot in a R object.
my_plot <- ggplot(data = tetrapods, mapping = aes(x = diet, y = lat, fill = diet)) + 
  geom_violin(scale = "width") +
  geom_boxplot(width=0.06, fill = "white", outlier.shape = NA) + #Adding a boxplot to the violinplot
  xlab("Diet") + ylab("Latitude (\u00B0)") +
  scale_fill_manual(values = colcajas) +
  scale_y_continuous(limits = c(-90, 90), breaks = c(-90,-50,0,50,90)) +
  scale_x_discrete(labels = c("Carnivore", "Herbivore", "Insectivore/Herbivore", "Piscivore")) +
  theme_classic() +
  theme(legend.position="none") 
```


5. Lastly, we can save our plots by `ggsave` function, among others.

```{r}
ggsave(file = "DV_ggplot1.svg", plot = my_plot, device = "svg", scale = 2, 
       width = 150, height = 100, units = "mm", dpi = 72, limitsize = TRUE)
```


6. Lastly, we should cite the packages we used, for example:

```{r}
citation("palaeoverse") 
```


## EXTRA!

7. We can change our plot font, using the `extrafont` package to import all the fonts.

```{r}
font_import()
fonts()[1:10] #Show the first 10 fonts
```

8. We also can customize the theme before calling the `ggplot` function. This is useful if we want to apply it to several plots. For more about theme configuration, see `?theme`

```{r}
Font1 <- "Palatino Linotype"
Size1 <- 13
Size2 <- 11
Map_col <- "#232B2B" 
Border_col <- "#FFFFFF"

theme <- theme(
  panel.background = element_blank(),
  plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm"),
  legend.justification= c(0,1),
  legend.position = c(0.05, 0.85),legend.title = element_blank(),
  legend.background = element_blank(),
  legend.key = element_rect(colour = NA, fill = NA),
  legend.key.size = unit(1.5, "lines"),
  legend.key.width = unit(1.5, "cm"),
  legend.text = element_text(
    size = Size2,
    face = "bold",
    family = Font1),
  panel.border = element_rect(colour = "black", fill = NA),
  axis.text.x = element_text(
    size = Size2,
    angle = 0,
    hjust = 0.5,
    family = Font1),
  axis.text.y = element_text(
    size = Size2,
    angle = 0,
    hjust = 0,
    family = Font1),
  axis.title.y = element_text(
    size = Size1,
    vjust = 5,
    family = Font1),
  axis.title.x = element_text(
    size = Size1,
    vjust = -1,
    family = Font1),
  aspect.ratio = 0.6
  )
```

Let's apply it to an example plot:

```{r}
#First, we will prepare the background of the plot.
data(GTS2020)
GTS2020 <- GTS2020[GTS2020$rank == "stage",]
GTS2020 <- GTS2020[GTS2020$min_ma >= 240 & GTS2020$max_ma <= 360,]

seq <- seq(1, 19, 2)
rect <- GTS2020[seq,]

#Then, we plot
ggplot() +
  geom_rect(
    data = rect,
    mapping = aes(
      xmin = min_ma,
      xmax = max_ma,
      ymin = 0,
      ymax = max(tetrapods_time$nrow) + 20
    ),
    linetype = 0,
    color = "grey90",
    alpha = 0.1
  )  +
  geom_line(data = tetrapods_time, aes(
    x = interval_mid_ma,
    y = nrow,
  ), linewidth = 1) +
  geom_point(data = tetrapods_time, aes(
    x = interval_mid_ma,
    y = nrow,
  ), size = 2.5) + 
  scale_x_reverse(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Time (Ma)", y = "Raw counts") +
  coord_geo(dat = list("stage", "epochs"), pos = list("b", "b"),
            xlim = c(355, 245), ylim = c(0,max(tetrapods_time$nrow)+20), 
            height = unit(1.5, "line"), size = 2.5,
            center_end_labels = T) +
  theme #this is the theme object from above
```

### References

ggplot2: Create Elegant Data Visualisations Using the Grammar of Graphics. https://cran.r-project.org/web/packages/ggplot2/index.html

R for Palaeobiologists: Workshop and Hackathon. https://hackathon.palaeoverse.org/
